Quantifying and Modeling Global Water Accessibility including Water Typology

### 2021-03-15 docs/manuscript/manuscript.tex

\section{Introduction}
In 2017, 29\% of the global population (2.2 billion people) did not use a safely managed drinking-water service, which is defined as water located on-premises, available, and free from contamination \citep{who2015drinkingwater}.   If you meet about 100 people, you can imagine that 29 of the people you meet do not have access to improved drinking water. Where this raises concerns about inequality problems related to the task and time variation 
to collect water can be inferred in the urban and rural areas \citep{cassivi2018access}. Therefore, there needs to be a
great motivation to bring improved drinking water sources to all people to improve their quality of life.

@@

% {\reversemarginpar\marginnote{\scriptsize\rd The introduction should end with a paragraph summarizing the contents and layout of the remainder of the manuscript. E.g. ``In the next section we discuss relevant prior research on water accessibility quantification. Next, we present the data and methods...''}} 
\section{Objectives}
The objective of this work is to  investigate associations between measures of water accessibility and access to modes of transit on the national scale.

In the next section, we discuss relevant prior research on water accessibility quantification. We are interested in what type of modes of
transportation are utilized to provide clean and drinking water to the community. Also, our objective is to investigate
the time it takes to transport water to the community in a global modeling context. So that we can develop and quantify
water accessibility in different countries to compare different trends. Also, we will be exploring the most efficient
and effective modes of transportation of accessible water. Therefore, the new models will be able to influence and
encourage policymakers in specific countries to improve water accessibility for all. Next, we present the data and methods of approaching the quantification of the water accessibility.

\section{Background}
Lack of water accessibility is a major crisis for many vulnerable people around the world. We have determined to seek literature reviews from the past 15 years to have the most up to date information because information on water accessibility is a topic that is constantly changing throughout the world due to different international, government, and even local policies. The literature review intends to investigate different journal articles on water accessibility on a global scale. Therefore we can conduct research to fill many major gaps to analyze the water accessibility predictor variables which could increase and enhance water accessibility on a global scale. We will always gain new insight and wisdom on information about water accessibility through the global scale as we investigate and determine literature reviews chronologically. 
\subsection{2000-2015}
Improving water accessibility has been a priority by many organizations, realizing it is the basic human right to have clean and accessible drinkable water in the past 15 years. From 2006 to 2016, overall water accessibility has been improving by utilizing the indicators of the past \citep{fuller2016tracking}. However, some researchers raise concerns on lack of clarity with the global water accessibility monitoring \citep{onda2012global, jia2016highresolution}. However, in an era of rapidly changing technologies, international and government policies, the older researches are not adequate. For an example of The eight Millennium Development Goals (MDGs) upgrading into The Sustainable Development Goals (SDGs). In light of the newest (SDGs) providing clean and accessible drinking water for all by 2030, researchers have devoted their studies to understand water accessibility at global and specific country levels. Several research and studies have focused global water accessibility ,and its progress \citep{sorensonSafeAccessSafe2011,bainAccountingWaterQuality2012,pickering2012freshwater,wssccWaterSanitationHygiene2014,  onda2012global, bartram2014global, onda2014country, who2015drinkingwater,fuller2016tracking, weiyuyuGlobalPerspectiveDrinkingwater2016, grahamAnalysisWaterCollection2016,overboOnplotDrinkingWater2016}. While there have been studies focusing on the local community in different parts of the country learning about their water accessibility and sanitation initiation \citep{jagals2006does, geereDomesticWaterCarrying2010, sima2013water, ho2014challenge, sakisakaIntroductionCommunityWater2015, jia2016highresolution}. Where majority raised concerns about the lack of monitoring sustainable water accessibility measurements for many people and some argue for a focus on many people's access to drinking water sources. Nonetheless, there is a lack of robust study on transportation to enhance water accessibility for all people by 2030. 
\subsection{2015-2021}
However, in the most recent years, there is a lack of information that needs to focus on people to receive accessible drinking water. The relation between water accessibility and the total time to retrieve drinking water is a concern and (SDGs) should emphasize this study \citep{cassivi2018access}. In the most recent years, there needs a greater understanding of the water accessibility indicators to enhance drinking water accessibility on a global scale for many vulnerable people. While research on water accessibility on a global scale has diminished due to an increase of water for many people \citep{fuller2016tracking}, but \citet{deshpande2020mapping} raises concerns that poorer community around the world are still not receiving clean and accessible drinking water. Several research and studies focus on a global scale to understand water accessibility and sanitation \citep{cassivi2018access, tortajada2018achieving, price2019difference,deshpande2020mapping}. Some research highlighted the coping cost in small community in India \citep{amit2019measuring} and researched on inequality with water accessibility and sanitation \citet{amit2019measuring,price2019difference, rawas2020comparing}. As many research journal articles focused on different aspects of water accessibility and sanitation in particular countries also brought greater insight and necessary to understand water accessibility variables \citep{rawas2020comparing, cassivi2021evaluating, martinez-santos2017determinants}. \citet{goal} emphasized providing safe and accessible drinking water for all on a global scale, which defines that we need a greater understanding of water accessibility and what water accessibility variables be investigated further. Therefore, we will be conducting robust research to understand water accessibility variables and predictors to improve water accessibility to bring clean and safe drinking water for all by 2030 as (SDGs) visioned.

@@

\subsection{Exploratory Data Analysis}
A correlation chart was used to assess strong correlations between the 28 numerical variables that may pertain to water accessibility \autoref{fig:cor-basic}. Households with limited water service has the strongest and positive relationship with households possessing a bicycle. Meanwhile, households using an improved water source has the strongest and negative relationship with households using an unimproved water source. This initial finding results indicated that the initial variables are logical and can give us insight into further understanding and the relationship between these variables.

\begin{figure}[h!]
  \centering
  \includegraphics[width=.7\textwidth]{correlation-basic}
  \caption{Correlation Graph of Numerical Variables}
  \label{fig:cor-basic}
\end{figure}
As we continued to study further upon the correlations of the variables, we used more advanced correlation graphs to determined their p-value, and to see a relationship between different categorical variables is shown in \autoref{fig:cor-plot}, which again indicates a greater detailed comparison between the variables. Households with an improved water source on the premises have a strong relationship with households with basic water service of .84 and a p-value of less than 0.001. Also, we learned and discovered that Households using an unimproved water source have a positive relationship with Households using unprotected well water, Households using an unprotected spring, and Households using surface water with a p-value of less than 0.001 as well. These relationships justify the necessity to research further if the particular water accessibility variables can increase the accessibility of particular water sources to reach universal access to water by 2030, as proposed in the SDGs.

@@

Additionally, we wanted to explore if the modes of transportation variables have any relationship with potential water variables. Therefore, we investigated by developing more correlation plots relating transportation variables with water accessibility variables which are shown in \autoref{fig:transporation-time}, \autoref{fig:transporation-water-one}, \autoref{fig:transporation-water-two}. There is a significant positive relationship between households possessing a private car and households with water on the premises as shown in \autoref{fig:transporation-time}. There is also is a significant negative relationship between households possessing a private car with households with water 30 minutes or less away round trip and households with water more than 30 minutes away round trip with a p-value of less than 0.001. This information is essential to investigate further why this is the global trend.

@@

\autoref{fig:transporation-water-one} shows that Households possessing a motorcycle has a positive correlation with Households using a tube well/borehole and Households using rainwater with a p-value of less than 0.01. Similar to the previous analysis, there is a relationship between a mode of transportation to retrieve potential water sources. \autoref{fig:transporation-water-two} also shows that households possessing an animal-drawn cart have a positive and strong relationship with households using unprotected well water with a p-value of less than 0.001. The results demonstrate that different modes of transportation factor into the accessibility of retrieving water sources on a global scale, therefore, there should be consideration of these variables to increase access to improved water sources. 
\begin{figure}[h!]
  \centering
  \includegraphics[width=.7\textwidth]{transportation-water-1}
  \caption{Advanced Correlation Graph of Modes of Transportation and Water Resources}
  \label{fig:transporation-water-one}
\end{figure}

@@

\subsection{Factor Analysis}

@@

NBClust statistic analysis developed and determined four distinct groups. The first cluster group is composed of countries near the equator or in the Sub-Sahara Regions. The analysis of the second group represents countries considered developed or recently industrialized. The third cluster group represents most of the countries in central Africa and may have similar factors as the first cluster group. The fourth cluster group represents most of the countries in the middle east or predominately non-African countries. There are great internally heterogeneous clusters for the 76 distinct countries. The cluster analysis' dendrogram is shown in \autoref{fig:dendrogram}, and the corresponding NBClust curve yielding the optimal four clusters is shown in \autoref{fig:optimal-clusters}, \autoref{fig:optimal-clusters-maj}, \autoref{fig:optimal}. \autoref{tab:water-cluster} shows the resulting four water accessibility country clusters. The cluster's water accessibility variables are summarized in \autoref{tab:summary-wa}. \autoref{fig:cluster-map} shows the cluster country membership of the water accessibility on a world map.  Through the optimal analysis of the clusters, which indicated the superior grouping of four cuts along with the relevant indicators.

@@

\section{Limitation}
%The problem of missing data is relatively common in almost all research and can have a significant effect on the conclusions that can be drawn from the data.

>>>>

\section{Introduction}
In 2017, 29\% of the global population (2.2 billion people) did not use a safely managed drinking water service, which is defined as water located on-premises, available, and free from contamination \citep{who2015drinkingwater}.   Where this raises concerns about inequality problems related to the task and time variation 
to collect water can be inferred in the urban and rural areas \citep{cassivi2018access}. Therefore, there needs to be a
great motivation to bring improved drinking water sources to all people to improve their quality of life.

@@

% {\reversemarginpar\marginnote{\scriptsize\rd The introduction should end with a paragraph summarizing the contents and layout of the remainder of the manuscript. E.g. ``In the next section we discuss relevant prior research on water accessibility quantification. Next, we present the data and methods...''}} 
%\section{Objectives}
The objective of this work is to  investigate associations between measures of water accessibility and access to modes of transit on the national scale. 
In the next section, we discuss relevant prior research on water accessibility quantification. We are interested in what type of modes of
transportation are utilized to provide clean and drinking water to the community. Also, our objective is to investigate
the time it takes to transport water to the community in a global modeling context. So that we can develop and quantify
water accessibility in different countries to compare different trends. Also, we will be exploring the most efficient
and effective modes of transportation of accessible water. Therefore, the new models will be able to influence and
encourage policymakers in specific countries to improve water accessibility for all. Next, we present the data and methods of approaching the quantification of the water accessibility.

\section{Background}
Lack of water accessibility is a major crisis for many vulnerable people around the world.  Information on water accessibility is a topic that is constantly changing throughout the world due to different international, government, and even local policies. The literature review intends to investigate different journal articles on water accessibility on a global scale. Therefore we can conduct research to fill many major gaps to analyze the water accessibility predictor variables which could increase and enhance water accessibility on a global scale. We will always gain new insight and wisdom on information about water accessibility through the global scale as we investigate and determine literature reviews chronologically.

\subsection{2000-2015}
Improving water accessibility has been a priority by many organizations, realizing it is the basic human right to have clean and accessible drinkable water in the past 15 years. From 2006 to 2016, overall water accessibility has been improving by utilizing the indicators of the past \citep{fuller2016tracking}. However, some researchers raise concerns on lack of clarity with the global water accessibility monitoring \citep{onda2012global, jia2016highresolution}. However, in an era of rapidly changing technologies, international and government policies, more relevant indicators are required. For an example of The eight Millennium Development Goals (MDGs) upgrading into The Sustainable Development Goals (SDGs). In light of the newest (SDGs) providing clean and accessible drinking water for all by 2030, researchers have devoted their studies to understand water accessibility at global and specific country levels. Several research and studies have focused global water accessibility ,and its progress \citep{sorensonSafeAccessSafe2011,bainAccountingWaterQuality2012,pickering2012freshwater,wssccWaterSanitationHygiene2014,  onda2012global, bartram2014global, onda2014country, who2015drinkingwater,fuller2016tracking, weiyuyuGlobalPerspectiveDrinkingwater2016, grahamAnalysisWaterCollection2016,overboOnplotDrinkingWater2016}. While there have been studies focusing on the local community in different parts of the country learning about their water accessibility and sanitation initiation \citep{jagals2006does, geereDomesticWaterCarrying2010, sima2013water, ho2014challenge, sakisakaIntroductionCommunityWater2015, jia2016highresolution}. Where majority raised concerns about the lack of monitoring sustainable water accessibility measurements for many people and some argue for a focus on many people's access to drinking water sources. Nonetheless, there is a lack of robust study on transportation to enhance water accessibility for all people by 2030.

\subsection{2015-current}
However, in the most recent years, there is a lack of information that needs to focus on people to receive accessible drinking water. The relation between water accessibility and the total time to retrieve drinking water is a concern and (SDGs) should emphasize this study \citep{cassivi2018access}. In the most recent years, there needs a greater understanding of the water accessibility indicators to enhance drinking water accessibility on a global scale for many vulnerable people. While research on water accessibility on a global scale has diminished due to an increase of water for many people \citep{fuller2016tracking}, but \citet{deshpande2020mapping} raises concerns that poorer community around the world are still not receiving clean and accessible drinking water. Several research and studies focus on a global scale to understand water accessibility and sanitation \citep{cassivi2018access, tortajada2018achieving, price2019difference,deshpande2020mapping}. Some research highlighted the coping cost in small community in India \citep{amit2019measuring} and researched on inequality with water accessibility and sanitation \citet{amit2019measuring,price2019difference, rawas2020comparing}. As many research journal articles focused on different aspects of water accessibility and sanitation in particular countries also brought greater insight and necessary to understand water accessibility variables \citep{rawas2020comparing, cassivi2021evaluating, martinez-santos2017determinants}. \citet{goal} emphasized providing safe and accessible drinking water for all on a global scale, which defines that we need a greater understanding of water accessibility and what water accessibility variables be investigated further. Therefore, we will be conducting robust research to understand water accessibility variables and predictors to improve water accessibility to bring clean and safe drinking water for all by 2030 as (SDGs) visioned.

@@

\subsection{Exploratory Data Analysis}
A correlation chart was used to assess strong correlations between the 28 numerical variables that may pertain to water accessibility \autoref{fig:cor-basic}. Countries with a high percentage of households with limited water service has the strongest and positive relationship with households possessing a bicycle. Meanwhile, countries with a high percentage of households using an improved water source has the strongest and negative relationship with countries with a high percentage of households using an unimproved water source. This initial finding results indicated that the initial variables are logical and can give us insight into further understanding and the relationship between these variables.

\begin{figure}[h!]
  \centering
  \includegraphics[width=.7\textwidth]{correlation-basic}
  \caption{Correlation Graph of Numerical Variables}
  \label{fig:cor-basic}
\end{figure}
As we continued to study further upon the correlations of the variables, we used more advanced correlation graphs to determined their p-value, and to see a relationship between different categorical variables is shown in \autoref{fig:cor-plot}, which again indicates a greater detailed comparison between the variables. Countries with a high percentage of households with an improved water source on the premises have a strong relationship with countries with a high percentage of households with basic water service of .84 and a p-value of less than 0.001. Also, we learned and discovered that countries with a high percentage of households using an unimproved water source have a positive relationship with countries with a high percentage of households using unprotected well water, countries with a high percentage of households using an unprotected spring, and countries with a high percentage of households using surface water with a p-value of less than 0.001 as well. The reason for the strong correlations is because the surveys has derivation from each of the responses. These relationships justify the necessity to research further if the particular water accessibility variables can increase the accessibility of particular water sources to reach universal access to water by 2030, as proposed in the SDGs.

@@

Additionally, we wanted to explore if the modes of transportation variables have any relationship with potential water variables. Therefore, we investigated by developing more correlation plots relating transportation variables with water accessibility variables which are shown in \autoref{fig:transporation-time}, \autoref{fig:transporation-water-one}, \autoref{fig:transporation-water-two}. There is a significant positive relationship between households possessing a private car and households with water on the premises as shown in \autoref{fig:transporation-time}.  There is also is a significant negative relationship between countries with a high percentage of households possessing a private car with countries with a high percentage of households with water 30 minutes or less away round trip and countries with a high percentage of households with water more than 30 minutes away round trip with a p-value of less than 0.001. This gives an inference that owning a private car could mean that the households are wealthy, and therefore they may have piped or close water sources nearby. This information is essential to investigate further why this is the global trend.

@@

\autoref{fig:transporation-water-one} shows that countries with a high percentage of households possessing a motorcycle has a positive correlation with countries with a high percentage of households using a tube well/borehole and countries with a high percentage of households using rainwater with a p-value of less than 0.01. Similar to the previous analysis, there is a relationship between a mode of transportation to retrieve potential water sources. \autoref{fig:transporation-water-two} also shows that countries with a high percentage of households possessing an animal-drawn cart have a positive and strong relationship with countries with a high percentage of households using unprotected well water with a p-value of less than 0.001. The analysis potentially demonstrates that different modes of transportation factor into the accessibility of retrieving water sources on a global scale, therefore, there should consideration of explanatory variables to enhance access to improved water sources. 
\begin{figure}[h!]
  \centering
  \includegraphics[width=.7\textwidth]{transportation-water-1}
  \caption{Advanced Correlation Graph of Modes of Transportation and Water Resources}
  \label{fig:transporation-water-one}
\end{figure}

@@

\subsection{Factor Analysis}
Before, we investigate the typologies or understand the clusters further, it is necessary to implement factor analysis to reduce the 17 water accessibility variables which is mentioned in \autoref{tab:summary-wa}. Factor analysis is a statistical method to explain variability among observed, correlated variables to decrease number of unobserved variables called factors.

@@

NBClust statistic analysis developed and determined four distinct groups of the four factors that were derived from water accessibility variables mentioned in \autoref{tab:summary-wa}. The first cluster group is composed of countries near the equator or in the Sub-Sahara Regions. The analysis of the second group represents countries considered developed or recently industrialized. The third cluster group represents most of the countries in central Africa and may have similar factors as the first cluster group. The fourth cluster group represents most of the countries in the middle east or predominately non-African countries. There are great internally heterogeneous clusters for the 76 distinct countries. The cluster analysis' dendrogram is shown in \autoref{fig:dendrogram}, and the corresponding NBClust curve yielding the optimal four clusters is shown in \autoref{fig:optimal-clusters}, \autoref{fig:optimal-clusters-maj}, \autoref{fig:optimal}. \autoref{tab:water-cluster} shows the resulting four water accessibility country clusters. The cluster's water accessibility variables are summarized in \autoref{tab:summary-wa}. \autoref{fig:cluster-map} shows the cluster country membership of the water accessibility on a world map.  Through the optimal analysis of the clusters, which indicated the superior grouping of four cuts along with the relevant indicators.

@@

\section{Limitations}
%The problem of missing data is relatively common in almost all research and can have a significant effect on the conclusions that can be drawn from the data.
